Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-05 paper.tex
handwaving

@@

Fig.~\ref{fig-ancestry-perf} is a useful yardstick, allowing us
to predict how long simulations should take for a wide range of
species. Taking a typical chromosome to be 1 Morgan in length,
these plots show, roughly, that simulating chromosome-length
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.
Simulating whole chromosomes for many species is very fast,
with 1000 samples of chromosome 1 for
\emph{Arabidopsis thaliana} taking less than a second, and a few
minutes for dogs and humans. However, the dependence on $\rho$
\emph{is} quadratic, and if $\rho$ is sufficiently large simulations
may not be feasible. For example, for
\emph{Drosophila melanogaster} chromosome 2L, with
$\rho\approx 3.9\times10^6$, Fig.~\ref{fig-ancestry-perf}
predicts that simulation would require around 177 hours. For such
large values of $\rho$  we recommend
users consider approximate simulations. Since \msprime\ does not currently
have efficient implementations of approximate coalescent with recombination
models, we recommend using SMC based methods such as \scrm\,
particularly for moderate sample sizes.
In practice, to predict the running time of a given
simulation in \msprime, we recommend that users
measure run time in a series of simulations with short genome lengths and
the desired sample size,
and then predict run time by fitting a quadratic curve to genome length
as in Fig.~\ref{fig-ancestry-perf}.
It is important to note that the quadratic curves in the two
panels of Fig.~\ref{fig-ancestry-perf} are different, and
predicting the run times of days-long simulations
using the timing of minutes-long runs is unlikely to be very accurate.

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of run time we can estimate when this occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
many (roughly half) of these lineages have at
least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
Then, since the rates of recombination and common ancestor events
are equal the time the maximum number of lineages is present,
we see that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

>>>>

% Peter L. Ralph
% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA

@@

Fig.~\ref{fig-ancestry-perf} is a useful yardstick, allowing us
to predict how long simulations should take for a wide range of
species. Taking a typical chromosome to be 1 Morgan in length,
these plots show, roughly, that simulating chromosome-length
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.
Simulating whole chromosomes for many species is very fast,
with 1000 samples of chromosome 1 for
\emph{Arabidopsis thaliana} taking less than a second, and a few
minutes for dogs and humans. However, the dependence on $\rho$
\emph{is} quadratic, and if $\rho$ is sufficiently large simulations
may not be feasible. For example,
the \emph{Drosophila melanogaster} chromosome 2L
is about 23.5Mb long and an average recombination rate of around $2.4 \times 10^{-8}$,
so $L \approx 0.57$,
and with $N_e = 1.7 \times 10^6$ \citep{li2006inferring},
$N_e L \approx 10^6$, so extrapolating the curve in Fig.~\ref{fig-ancestry-perf}B
predicts that simulation would require around 177 hours.
For such large values of $\rho$  we recommend
users consider approximate simulations. Since \msprime\ does not currently
have efficient implementations of approximate coalescent with recombination
models, we recommend using SMC based methods such as \scrm\,
particularly for moderate sample sizes.
In practice, to predict the running time of a given
simulation in \msprime, we recommend that users
measure run time in a series of simulations with short genome lengths and
the desired sample size,
and then predict run time by fitting a quadratic curve to genome length
as in Fig.~\ref{fig-ancestry-perf}.
It is important to note that the quadratic curves in the two
panels of Fig.~\ref{fig-ancestry-perf} are different, and
predicting the run times of days-long simulations
using the timing of minutes-long runs is unlikely to be very accurate.

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of run time we can estimate when this occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
At the maximum, coalescence and recombination rates are equal,
so a typical segment of ancestry will spend roughly half its time
in a lineage with at least one other such segment---and,
since such lineages carry at least two segments,
at most one-third of the lineages carry long trapped segments of ancesry.
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
this implies that $\ell \approx L / 6$.
Setting the rates of recombination and common ancestor events
to be equal and solving for $M$,
we find that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 6$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.
